---
title: "decomp_stats_figures"
output: pdf_document
date: "2026-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This script is a general guide for making publication-quality visualizations using QIIME2-generated outputs

################################################################################
# Research Questions
1. Does microbial community composition change over accumulated degree days?
2. Does facility influence microbial diversity or community composition?
3. Do soil and skin samples differ in their microbial diversity and community composition?

- Your research question doesn’t need to be included in your R Markdown, but you should have a data analysis plan. It’s easy to go down rabbit holes if you don’t stick to analyses that directly address your questions.
################################################################################


################################################################################
# Installing and loading packages
################################################################################

# Install packages
```{r}
# Each of the packages we will use are CRAN packages
packages_to_install <- c("tidyverse", "ggplot2","readr","ggpubr","vegan","scales","lme4", "lmerTest","pbkrtest","emmeans")

install.packages(packages_to_install)
```

# Load packages
```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(ggpubr)
library(vegan)
library(scales)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(emmeans)
```
################################################################################
# Metadata
################################################################################

# Import and clean metadata if needed
```{r}
metadata <- read_tsv("../03_metadata/metadata.txt")
```


################################################################################
# Alpha diversity
################################################################################

# Import alpha diversity data from QIIME2 output
```{r}
# Alpha diversity metrics
shannon <- read_tsv("alpha_div/shannon.tsv")
evenness <- read_tsv("alpha_div/evenness.tsv")
observed_features <- read_tsv("alpha_div/observed_features.tsv")
faiths_pd <- read_tsv("alpha_div/faith_pd.tsv")
```
# Tidy metrics and add to metadata
```{r}
# Shannon 
shannon <- shannon %>%
  rename(sample_name = 1)  #rename 1st column to to match metadata

# Evenness
evenness <- evenness %>% 
  rename(sample_name = 1)  #rename 1st column to to match metadata

# Observed Features
observed_features <- observed_features %>% 
  rename(sample_name = 1)  #rename 1st column to to match metadata

# Faiths PD
faiths_pd <- faiths_pd %>% 
  rename(sample_name = 1)  #rename 1st column to to match metadata

# Add to metadata
metadata <- metadata %>%
  left_join(shannon, by = "sample_name") %>% 
  left_join(evenness, by = "sample_name") %>% 
  left_join(observed_features, by = "sample_name") %>% 
  left_join(faiths_pd, by = "sample_name")
  

```

* If you look at your metadata you will see that 4 columns (variables) were added this is the metrics from the alpha diversity added as columns to the end of your metadata. 

* You'll also notice there are some samples that have NA's for the metrics, these could have been filtered out during our Qiime2 analysis. It is good practice to check these NA's to make sure you aren't missing data. Then you can filter them out. 

# Tidy metadata with metrics 
```{r}
metadata <- metadata %>% 
  drop_na(shannon_entropy, pielou_evenness, observed_features, faith_pd)
```

* Now you'll notice that we lost some columns, but there are no NA's in our metrics columns, now we are ready to do statistical analysis. 

# Linear Mixed Effects Models
* When building linear mixed-effects models, it’s best to start with the simplest model that adequately fits your data. I usually begin by including the variable that I expect to have the largest effect—or the one that might obscure the treatment signal—so we can account for it before testing our main hypothesis. In this case we will start with add_0 since it is our time factor. 

## Time (add_0c) (question 1)

Shannon
```{r}
# Build model
model_add0_shannon <- lmer(shannon_entropy ~ add_0c + (1|host_subject_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_add0_shannon)
# look at estimated variance components
summary(model_add0_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0_shannon)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```
For this first metric we are going to walk through each command, what they do, and what you should look for. 

Model: model_add0_shannon <- lmer(shannon_entropy ~ add_0c + (1|host_subject_id), data = metadata)
  - Model type: We are fitting a linear mixed-effects model using the lmer() function from the lme4 package. Mixed-effects models are useful when data points are not independent, such as repeated measurements from the same subject.
  - Response variable: shannon_entropy — the dependent variable we want to explain (Shannon diversity in this case).
  - Fixed effect: add_0c — the predictor we are testing, treated as a fixed effect. The model estimates how changes in add_0c influence Shannon diversity across all subjects.
  - Random effect:
(1|host_subject_id) — a random intercept for each subject. This accounts for repeated measurements between subjects.
  - Data: data = metadata

1. plot(model_add0_shannon) Purpose: Checks for non-linearity and heteroscedasticity (unequal variance).
  - What to look for:Residuals should be roughly randomly scattered around 0.No obvious       patterns, trends, or funnels.
  - Red flag:Curved or fan-shaped pattern, the model may not fit well, or variance may not be constant. You may need to consider a different model for your data. 

2. summary(model_add0_shannon)
  - This tells us how Shannon diversity changes with add_0c, while accounting for differences between individual hosts. The fixed effects tell us the average effect across all hosts, and the random effects tell us how much subjects differ from each other.
  - Look for: Fixed effect estimates make sense (sign, magnitude), standard errors are reasonable, and random effects show meaningful variation between subjects.
  - Red flags: Very large or tiny standard errors, fixed effect estimates that don’t align with expectations, random effects variance near zero, or unusually high residual variance suggesting poor model fit.

3. anova(model_add0_shannon)
  - Performs a Type III or sequential ANOVA on the model, testing the significance of fixed effects (e.g., add_0c) in explaining the response.
  - Look for: F-values and p-values indicate whether each fixed effect contributes significantly to explaining variation in shannon_entropy.
  - Red flags: Very high or very low F-values with inconsistent p-values, or effects that don’t match biological expectations; may indicate model misspecification, collinearity, or insufficient data.

Evenness
```{r}
# Build model
model_add0_evenness <- lmer(pielou_evenness ~ add_0c + (1|host_subject_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_add0_evenness)
# look at estimated variance components
summary(model_add0_evenness)
# residual diagnostic plot, check assumptions
anova(model_add0_evenness)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```


Observed Features
```{r}
# Build model
model_add0_obfeatures <- lmer(observed_features ~ add_0c + (1|host_subject_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_add0_obfeatures)
# look at estimated variance components
summary(model_add0_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_add0_obfeatures)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```


Faiths PD
```{r}
# Build model
model_add0_faithpd <- lmer(faith_pd ~ add_0c + (1|host_subject_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_add0_faithpd)
# look at estimated variance components
summary(model_add0_faithpd)
# residual diagnostic plot, check assumptions
anova(model_add0_faithpd)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```

* If you remember when we were looking at our qiime2 outputs we could see that something was going at at the begining of the study vs the end of the study so lets add a metadata column that splits the metadata into 3 sections. When looking at the metadata we can see the highest add_0c was 450. (This is just to see what is going on, once you take a look then you can decided based on your data a more appropriate way to look at time)

## add_0c split into 3

Shannon 
```{r}
# Add metadata column 
metadata <- metadata %>%
  mutate(add0c_split = cut(
    add_0c,
    breaks = 3,
    labels = c("0-150", "150-300", "300-450"),
    include.lowest = TRUE
  ))

# Stats 
# Build model
model_add0split_shannon <- lmer(shannon_entropy ~ add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0split_shannon)
# look at estimated variance components
summary(model_add0split_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0split_shannon)
# pairwise comparisons
emmeans(model_add0split_shannon, pairwise ~ add0c_split)
```
* From our emmeans we aren't seeing anything in shannon diversity specifically. 

On your own you can try splitting add_0c in two groups. 
```{r}
metadata <- metadata %>%
  mutate(add0c_split2 = cut(
    add_0c,
    breaks = 2,
    labels = c("0-225", "225-425"),
    include.lowest = TRUE
  ))

# Stats shannon
# Build model
model_add0split2_shannon <- lmer(shannon_entropy ~ add0c_split2 + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0split2_shannon)
# look at estimated variance components
summary(model_add0split2_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0split2_shannon)
# pairwise comparisons
emmeans(model_add0split2_shannon, pairwise ~ add0c_split2)
```

# add_0c figure (voltility plot)
```{r}

```


## add0c + facility (question 2)
To make it easier to demonstrate time I am going to continue the analysis with the 3 group add_0c split. 
In this section, we will test whether microbial community composition changes differently across facilities over accumulated degree days (add_0c). To do this, we included facility as an interaction with add_0c in the model. This allows us to evaluate whether the trajectory of microbial community change over time differs between facilities, rather than simply testing for overall differences between facilities. By modeling the interaction, we can determine whether microbial communities diverge from their baseline at different rates depending on the facility where the samples were collected.
Shannon
```{r}
# Build model
model_add0_facility_shannon <- lmer(shannon_entropy ~ facility*add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_shannon)
# look at estimated variance components
summary(model_add0_facility_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_shannon)
# pairwise comparisons
emmeans(model_add0_facility_shannon, pairwise ~ facility|add0c_split)
```
Really quick lets just look when we don't split add_0c
```{r}
# Build model
model_add0_facility_shannon2 <- lmer(shannon_entropy ~ facility*add_0c + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_shannon2)
# look at estimated variance components
summary(model_add0_facility_shannon2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_shannon2)
# pairwise comparisons
emmeans(model_add0_facility_shannon2, pairwise ~ facility)
```

Evenness
```{r}
# Build model
model_add0_facility_evenness <- lmer(pielou_evenness ~ facility*add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_evenness)
# look at estimated variance components
summary(model_add0_facility_evenness)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_evenness)
# pairwise comparisons
emmeans(model_add0_facility_evenness, pairwise ~ facility|add0c_split)
```

Observed Features
```{r}
# Build model
model_add0_facility_obfeatures <- lmer(observed_features ~ facility*add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_obfeatures)
# look at estimated variance components
summary(model_add0_facility_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_obfeatures)
# pairwise comparisons
emmeans(model_add0_facility_obfeatures, pairwise ~ facility|add0c_split)
```

Faith's PD
```{r}
# Build model
model_add0_facility_faithpd <- lmer(faith_pd ~ facility*add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_faithpd)
# look at estimated variance components
summary(model_add0_facility_faithpd)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_faithpd)
# pairwise comparisons
emmeans(model_add0_facility_faithpd, pairwise ~ facility|add0c_split)
```
- We can see that facility is significant at earlier timepoints so when we look at sample_type we will include it in our model. 

# Facilty figure ajusted for time (box plot) (3 timepoint plot)




# Facility figure over time (voltility plot)


## add_0c + facility + sample_type (question 3)
We know that facility varies across accumulated degree days (add_0c), but it is not the primary variable addressing our research question in this section. Therefore, we include facility in the model as an additive term rather than as an interaction term. Instead, we include sample_type as an interaction with add_0c, because we want to test whether microbial communities from soil and skin change differently over time. We previously examined the effect of facility in an earlier section, so here it is included only as a covariate to account for background variation. Specifically, we are testing whether the time trajectories differ between soil and skin samples. If we instead wanted to compare soil versus skin while simply accounting for time, rather than testing different time trajectories, add_0c would be included as an additive term rather than in an interaction.

Shannon 
```{r}
# Build model
model_add0_facility_sampletype_shannon <- lmer(shannon_entropy ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_shannon)
# look at estimated variance components
summary(model_add0_facility_sampletype_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_shannon)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_shannon, pairwise ~ sample_type|add0c_split)
```

Evenness
```{r}
# Build model
model_add0_facility_sampletype_evenness <- lmer(pielou_evenness ~ add0c_split*sample_type + facility + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_evenness)
# look at estimated variance components
summary(model_add0_facility_sampletype_evenness)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_evenness)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_evenness, pairwise ~ sample_type|add0c_split)
```

Observed features
```{r}
# Build model
model_add0_facility_sampletype_obfeatures <- lmer(observed_features ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_obfeatures)
# look at estimated variance components
summary(model_add0_facility_sampletype_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_obfeatures)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_obfeatures, pairwise ~ sample_type|add0c_split)
```

Faith's PD
```{r}
# Build model
model_add0_facility_sampletype_faithpd <- lmer(faith_pd ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_faithpd)
# look at estimated variance components
summary(model_add0_facility_sampletype_faithpd)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_faithpd)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_faithpd, pairwise ~ sample_type|add0c_split)
```

## Sample type figure
```{r}

```




################################################################################
# Beta diversity
################################################################################

# Load in beta diversity metrics and tidy to add to metadata
(remember this is already our PCOA results so this is already our ordinations). I make separate metadata's for each beta diversity metric, another option would be to rename each PC ex. PC1J for jaccard and add PC1-PC3 for each metric to the metadata. I find it easier to keep everything seperate when the metrics have different metadata. 
```{r}
# Read in beta div metrics 
bray_curtis <- read_tsv("beta_div/bray_curtis.txt")
jaccard <- read_tsv("beta_div/jaccard.txt")
uw_unifrac <- read_tsv("beta_div/unweighted_unifrac.txt")
w_unifrac <- read_tsv("beta_div/weighted_unifrac.txt")

# Clean metrics
# bray curtis
# get variance
bray_curtis_var <- bray_curtis %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# clean ordination
bray_curtis <- bray_curtis %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_bc <- bray_curtis %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))


# Jaccard
# get variance
jaccard_var <- jaccard %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Clean ordinace
jaccard <- jaccard %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_jaccard <- jaccard %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Unweighted Unifrac
uw_unifrac <- uw_unifrac %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_uwunifrac <- uw_unifrac %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Weighted Unifrac
w_unifrac <- w_unifrac %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_wunifrac <- w_unifrac %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

```
- Now that we've done some data tidying we can finally get to stats for beta diversity, I am going to keep using the add0c split for demonstration purposes. 

- There are two common approaches for performing statistical analyses on beta diversity. One approach is first distances, which we discussed last week. If you want to run a linear mixed-effects model (LME) in R using the first-distances output, you can unzip the first-distances.qza artifact and extract the data file. After importing this file into R, you can use it in your model. In this case, instead of using a PCoA axis (e.g., PC1) as the dependent variable, you would use Distance as the dependent variable in the LME.

-  For the sake of time, we will skip ahead to question 3 and build a model examining sample type. In this model, we account for facility as a covariate and evaluate how beta diversity changes over time. Specifically, we include an interaction between accumulated degree days (add_0c) and sample type to test whether the trajectory of beta diversity over time differs between soil and skin samples. 
## add0c_split + facility + sample_type (question 3)
- If you think back to the PCoA plots, we saw three axes: PC1, PC2, and PC3. We will test each of these axes to determine whether the variance along that axis is explained by our variables. In many microbiome datasets, time often explains a large portion of the variance on PC1, although this is not always the case. 

# Bray Curtis

PC1
```{r}
# Build model
model_add0_facility_sampletype_bc1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc1)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc1, pairwise ~ sample_type|add0c_split)
```

PC2
```{r}
# Build model
model_add0_facility_sampletype_bc2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc2)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc2, pairwise ~ sample_type|add0c_split)
```


PC3
```{r}
# Build model
model_add0_facility_sampletype_bc3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc3)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc3, pairwise ~ sample_type|add0c_split)
```

# Jaccard

PC1
```{r}
# Build model
model_add0_facility_sampletype_j1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j1)
# look at estimated variance components
summary(model_add0_facility_sampletype_j1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j1, pairwise ~ sample_type|add0c_split)
```


PC2
```{r}
# Build model
model_add0_facility_sampletype_j2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j2)
# look at estimated variance components
summary(model_add0_facility_sampletype_j2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j2, pairwise ~ sample_type|add0c_split)
```


PC3
```{r}
# Build model
model_add0_facility_sampletype_j3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j3)
# look at estimated variance components
summary(model_add0_facility_sampletype_j3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j3, pairwise ~ sample_type|add0c_split)
```
# Unweighted Unifrac

PC1
```{r}
# Build model
model_add0_facility_sampletype_uwunifrac1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac1)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac1, pairwise ~ sample_type|add0c_split)
```
PC2
```{r}
# Build model
model_add0_facility_sampletype_uwunifrac2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac2)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac2, pairwise ~ sample_type|add0c_split)
```
PC3
```{r}
# Build model
model_add0_facility_sampletype_uwunifrac3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac3)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac3, pairwise ~ sample_type|add0c_split)
```
# Weighted Unifrac

PC1
```{r}
# Build model
model_add0_facility_sampletype_wunifrac1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac1)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac1, pairwise ~ sample_type|add0c_split)
```
PC2
```{r}
# Build model
model_add0_facility_sampletype_wunifrac2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac2)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac2, pairwise ~ sample_type|add0c_split)
```
PC3
```{r}
# Build model
model_add0_facility_sampletype_wunifrac3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac3)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac3, pairwise ~ sample_type|add0c_split)
```

# Figure

```{r}
# Bray Curtis

# Get variance
prop_row <- bray_curtis %>%
  filter(Eigvals == "Proportion explained")

prop_values <- prop_row %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t")

bray_curtis_var <- prop_values %>%
  mutate(across(starts_with("PC"), as.numeric)) %>%
  mutate(across(starts_with("PC"), ~ .x * 100))

bray_curtis_var %>%
  select(PC1, PC2, PC3)

# Plot
bray_curtis_var <- bray_curtis$data$ProportionExplained * 100

age_bray_curtis <- ggplot(metadata_bray_curtis, aes(x = PC1, y = PC2, color = age_years)) + 
  geom_point(size = 3) +
  theme_bw() + 
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(bray_curtis_var[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(bray_curtis_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("lt24" = "salmon", "gt24" = "skyblue"),
    labels = c("lt24" = "Younger than 25 months", "gt24" = "Older than 24 months"),
    name = "Age"
  ) +
  labs(title = "Bray-Curtis PCoA") +
annotate("text",
           x = x_max * 1.2,
           y = y_max * 1.2,
           label = "p = 0.0078",
           hjust = 1, vjust = 1,
           size = 5, fontface = "italic") +
  coord_cartesian(clip = "off")

age_bray_curtis
```

