---
title: "decomp_stats_figures"
output: pdf_document
date: "2026-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This script is a general guide for making publication-quality visualizations using QIIME2-generated outputs

################################################################################
# Research Questions
1. Does microbial community composition change over accumulated degree days?
2. Does facility influence microbial diversity or community composition?
3. Do soil and skin samples differ in their microbial diversity and community composition?

- Your research question doesn’t need to be included in your R Markdown, but you should have a data analysis plan. It’s easy to go down rabbit holes if you don’t stick to analyses that directly address your questions.
################################################################################


################################################################################
# Installing and loading packages
################################################################################

# Install packages
- If you have already installed the packages then you don't need to rerun this. 
```{r}
# Each of the packages we will use are CRAN packages
packages_to_install <- c("tidyverse", "ggplot2","readr","ggpubr","vegan","scales","lme4", "lmerTest","pbkrtest","emmeans", "rstatix", "forcats", "scales")

install.packages(packages_to_install)
```

# Load packages
```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(ggpubr)
library(vegan)
library(scales)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(emmeans)
library(rstatix)
library(forcats)
library(scales) 
```
################################################################################
# Metadata
################################################################################

# Import and clean metadata if needed
```{r}
metadata <- read_tsv("../03_metadata/metadata.txt")
```


################################################################################
# Alpha diversity
################################################################################

# Import alpha diversity data from QIIME2 output
```{r}
# Alpha diversity metrics
shannon <- read_tsv("alpha_div/shannon.tsv")
evenness <- read_tsv("alpha_div/evenness.tsv")
observed_features <- read_tsv("alpha_div/observed_features.tsv")
faiths_pd <- read_tsv("alpha_div/faith_pd.tsv")
```
# Tidy metrics and add to metadata
```{r}
# Shannon 
shannon <- shannon %>%
  rename(sample_name = 1)  #rename 1st column to to match metadata

# Evenness
evenness <- evenness %>% 
  rename(sample_name = 1)  #rename 1st column to to match metadata

# Observed Features
observed_features <- observed_features %>% 
  rename(sample_name = 1)  #rename 1st column to to match metadata

# Faiths PD
faiths_pd <- faiths_pd %>% 
  rename(sample_name = 1)  #rename 1st column to to match metadata

# Add to metadata
metadata <- metadata %>%
  left_join(shannon, by = "sample_name") %>% 
  left_join(evenness, by = "sample_name") %>% 
  left_join(observed_features, by = "sample_name") %>% 
  left_join(faiths_pd, by = "sample_name")
  

```

* If you look at your metadata you will see that 4 columns (variables) were added this is the metrics from the alpha diversity added as columns to the end of your metadata. 

* You'll also notice there are some samples that have NA's for the metrics, these samples could have been filtered out during our Qiime2 analysis and thus did not generate an alpha diversity metrics during the core-metrics command (because rarefaction will filter out samples less than the read depth you give it!). 

It is good practice to check these NA's to make sure you aren't missing data. Then you can filter the NAs out. 

# Tidy metadata with metrics 
```{r}
metadata <- metadata %>% 
  drop_na(shannon_entropy, pielou_evenness, observed_features, faith_pd)
```

* Now you'll notice that we lost some columns, but there are no NA's in our metrics columns, now we are ready to do statistical analysis. 

# Linear Mixed Effects Models
* When building linear mixed-effects models, it’s best to start with the simplest model that adequately fits your data. I usually begin by including the variable that I expect to have the largest effect—or the one that might obscure the treatment signal—so we can account for it before testing our main hypothesis. In this case we will start with add_0 since it is our time factor. 


## Body Site (body_site) (question x)

## Time (add_0c) (question 1)

Shannon
```{r}
# Build model
model_add0_shannon <- lmer(shannon_entropy ~ add_0c + (1|host_subject_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_add0_shannon)
# look at estimated variance components
summary(model_add0_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0_shannon)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```
For this first metric we are going to walk through each command, what they do, and what you should look for. 

Model: model_add0_shannon <- lmer(shannon_entropy ~ add_0c + (1|host_subject_id), data = metadata)
  - Model type: We are fitting a linear mixed-effects model using the lmer() function from the lme4 package. Mixed-effects models are useful when data points are not independent, such as repeated measurements from the same subject.
  - Response variable: shannon_entropy — the dependent variable we want to explain (Shannon diversity in this case).
  - Fixed effect: add_0c — the predictor we are testing, treated as a fixed effect. The model estimates how changes in add_0c influence Shannon diversity across all subjects.
  - Random effect:
(1|host_subject_id) — a random intercept for each subject. This accounts for repeated measurements between subjects.
  - Data: data = metadata

1. plot(model_add0_shannon) Purpose: Checks for non-linearity and heteroscedasticity (unequal variance).
  - What to look for:Residuals should be roughly randomly scattered around 0.No obvious patterns, trends, or funnels.
  - Red flag:Curved or fan-shaped pattern, the model may not fit well, or variance may not be constant. You may need to consider a different model for your data. 

2. summary(model_add0_shannon)
  - This tells us how Shannon diversity changes with add_0c, while accounting for differences between individual hosts. The fixed effects tell us the average effect across all hosts, and the random effects tell us how much subjects differ from each other.
  - Look for: Fixed effect estimates make sense (sign, magnitude), standard errors are reasonable, and random effects show meaningful variation between subjects.
  - Red flags: Very large or tiny standard errors, fixed effect estimates that don’t align with expectations, random effects variance near zero, or unusually high residual variance suggesting poor model fit.

3. anova(model_add0_shannon)
  - Performs a Type III or sequential ANOVA on the model, testing the significance of fixed effects (e.g., add_0c) in explaining the response.
  - Look for: F-values and p-values indicate whether each fixed effect contributes significantly to explaining variation in shannon_entropy.
  - Red flags: Very high or very low F-values with inconsistent p-values, or effects that don’t match biological expectations; may indicate model misspecification, collinearity, or insufficient data.

Evenness
```{r}
# Build model
model_add0_evenness <- lmer(pielou_evenness ~ add_0c + (1|host_subject_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_add0_evenness)
# look at estimated variance components
summary(model_add0_evenness)
# residual diagnostic plot, check assumptions
anova(model_add0_evenness)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```


Observed Features
```{r}
# Build model
model_add0_obfeatures <- lmer(observed_features ~ add_0c + (1|host_subject_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_add0_obfeatures)
# look at estimated variance components
summary(model_add0_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_add0_obfeatures)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```


Faiths PD
```{r}
# Build model
model_add0_faithpd <- lmer(faith_pd ~ add_0c + (1|host_subject_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_add0_faithpd)
# look at estimated variance components
summary(model_add0_faithpd)
# residual diagnostic plot, check assumptions
anova(model_add0_faithpd)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```

* If you remember when we were looking at our qiime2 outputs we could see that something was going at at the begining of the study vs the end of the study so lets add a metadata column that splits the metadata into 3 sections. When looking at the metadata we can see the highest add_0c was 450. (This is just to see what is going on, once you take a look then you can decided based on your data a more appropriate way to look at time)

## add_0c split into 3

Shannon 
```{r}
# Add metadata column 
metadata <- metadata %>%
  mutate(add0c_split = cut(
    add_0c,
    breaks = 3,
    labels = c("0-150", "150-300", "300-450"),
    include.lowest = TRUE
  ))

# Stats 
# Build model
model_add0split_shannon <- lmer(shannon_entropy ~ add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0split_shannon)
# look at estimated variance components
summary(model_add0split_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0split_shannon)
# pairwise comparisons
emmeans(model_add0split_shannon, pairwise ~ add0c_split)
```
* From our emmeans we aren't seeing anything in shannon diversity specifically. 

On your own you can try splitting add_0c in two groups. 
```{r}
metadata <- metadata %>%
  mutate(add0c_split2 = cut(
    add_0c,
    breaks = 2,
    labels = c("0-225", "225-425"),
    include.lowest = TRUE
  ))

# Stats shannon
# Build model
model_add0split2_shannon <- lmer(shannon_entropy ~ add0c_split2 + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0split2_shannon)
# look at estimated variance components
summary(model_add0split2_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0split2_shannon)
# pairwise comparisons
emmeans(model_add0split2_shannon, pairwise ~ add0c_split2)
```

# add_0c figure (voltility plot)

```{r}
# Shannon
add0c_shannon <- ggplot(metadata, aes(x = add_0c, y = shannon_entropy, linetype = sample_type)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Accumulated Degree Days (ADD)",
    y = "Shannon Entropy",
    linetype = "Sample Type",
    title = "Shannon"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

add0c_shannon

# Evennes
add0c_evenness <- ggplot(metadata, aes(x = add_0c, y = pielou_evenness, linetype = sample_type)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Accumulated Degree Days (ADD)",
    y = "Pielou's Evenness",
    linetype = "Sample Type",
    title = "Evenness"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

add0c_evenness

# Observed Features
add0c_obfeatures <- ggplot(metadata, aes(x = add_0c, y = observed_features, linetype = sample_type)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Accumulated Degree Days (ADD)",
    y = "Observed Features",
    linetype = "Sample Type",
    title = "Observed Features"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

add0c_obfeatures

# Faith's PD
add0c_faithpd <- ggplot(metadata, aes(x = add_0c, y = faith_pd, linetype = sample_type)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Accumulated Degree Days (ADD)",
    y = "Faith's PD",
    linetype = "Sample Type",
    title = "Faith's PD"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

add0c_faithpd

# Combine the plots
alpha_div_addOc <- ggarrange(add0c_shannon, add0c_evenness,
                                  add0c_obfeatures, add0c_faithpd,
                                  ncol = 2, nrow = 2,
                                  labels = c("A","B","C","D"),
                                  common.legend = TRUE,
                                  legend = "bottom")
alpha_div_addOc

# Save the figure
ggsave(alpha_div_addOc, filename = "../05_figures/alpha_div_addOc.jpeg", device = "jpeg", dpi = 300, width = 12, height = 9)

```



# add_0c 3 group figure

```{r}
model_add0split_shannon <- lmer(shannon_entropy ~ add0c_split + (1|host_subject_id), data = metadata)

pvals <- pairs(emmeans(model_add0split_shannon, ~ add0c_split)) |> as.data.frame()

# make table for stat_pvalue_manual with rounded p-values
pvals_table <- data.frame(
  group1 = c("0-150","0-150","150-300"),
  group2 = c("150-300","300-450","300-450"),
  p = round(pvals$p.value, 3),           # round to 3 decimals
  y.position = max(metadata$shannon_entropy) + c(0.1,0.3,0.5)
)

# plot
add_split_shannon <- ggplot(metadata, aes(x = add0c_split, y = shannon_entropy)) +
  geom_boxplot() +
  stat_pvalue_manual(pvals_table, label = "p") +
  theme_bw() +
  theme(
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"))+
  labs(x = "Accumulated Degree Day 3 Groups",
       y = "Shannon Entropy",
       title = "Shannon Diversity for ADD Split into 3 Groups")

add_split_shannon

# Save the figure
ggsave(add_split_shannon, filename = "../05_figures/add_split_shannon.jpeg", device = "jpeg", dpi = 300, width = 8, height = 9)
```



## add0c + facility (question 2)
To make it easier to demonstrate time I am going to continue the analysis with the 3 group add_0c split. 
In this section, we will test whether microbial community composition changes differently across facilities over accumulated degree days (add_0c). To do this, we included facility as an interaction with add_0c in the model. This allows us to evaluate whether the trajectory of microbial community change over time differs between facilities, rather than simply testing for overall differences between facilities. By modeling the interaction, we can determine whether microbial communities diverge from their baseline at different rates depending on the facility where the samples were collected.
Shannon
```{r}
# Build model
model_add0_facility_shannon <- lmer(shannon_entropy ~ facility*add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_shannon)
# look at estimated variance components
summary(model_add0_facility_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_shannon)
# pairwise comparisons
emmeans(model_add0_facility_shannon, pairwise ~ facility|add0c_split)
```
Really quick lets just look when we don't split add_0c
```{r}
# Build model
model_add0_facility_shannon2 <- lmer(shannon_entropy ~ facility*add_0c + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_shannon2)
# look at estimated variance components
summary(model_add0_facility_shannon2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_shannon2)
# pairwise comparisons
emmeans(model_add0_facility_shannon2, pairwise ~ facility)
```

Evenness
```{r}
# Build model
model_add0_facility_evenness <- lmer(pielou_evenness ~ facility*add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_evenness)
# look at estimated variance components
summary(model_add0_facility_evenness)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_evenness)
# pairwise comparisons
emmeans(model_add0_facility_evenness, pairwise ~ facility|add0c_split)
```

Observed Features
```{r}
# Build model
model_add0_facility_obfeatures <- lmer(observed_features ~ facility*add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_obfeatures)
# look at estimated variance components
summary(model_add0_facility_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_obfeatures)
# pairwise comparisons
emmeans(model_add0_facility_obfeatures, pairwise ~ facility|add0c_split)
```

Faith's PD
```{r}
# Build model
model_add0_facility_faithpd <- lmer(faith_pd ~ facility*add0c_split + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_faithpd)
# look at estimated variance components
summary(model_add0_facility_faithpd)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_faithpd)
# pairwise comparisons
emmeans(model_add0_facility_faithpd, pairwise ~ facility|add0c_split)
```
- We can see that facility is significant at earlier timepoints so when we look at sample_type we will include it in our model. 

# Facilty figure ajusted for time (box plot) (3 timepoint box plot)
```{r}
model_add0_facility_faithpd <- lmer(faith_pd ~ facility*add0c_split + (1|host_subject_id), data = metadata)  
pvals <- pairs(
  emmeans(model_add0_facility_faithpd, ~ facility | add0c_split)
) |> as.data.frame()
pvals_table <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  dplyr::mutate(
    p = round(p.value, 4),
    y.position = max(metadata$faith_pd) + 0.1
  )
pvals_table

#only plot the significant associations
pvals_table_significant <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  filter(p.value < 0.05) %>%                     
  mutate(p = round(p.value, 4))
pvals_table_significant

# fix overlapping significance bars
pvals_table_significant <- pvals_table_significant %>%
  add_y_position(
    data = metadata,
    formula = faith_pd ~ facility,
    fun = "max",
    step.increase = 0.12
  )
pvals_table_significant


# Plot Faiths PD per facility by the grouped ADDs, add only significant comparisons
add_split_faithpd_facility <- ggplot(metadata,
       aes(x = facility, y = faith_pd, fill = facility)) +
  geom_boxplot() +
  facet_wrap(~ add0c_split) +
  stat_pvalue_manual(
    pvals_table_significant,
    xmin = "group1",
    xmax = "group2",
    label = "p",
    tip.length = 0.01,
    size = 3, 
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = c(
    "FIRS" = "#ff9896",
    "STAFS" = "#8c6d31",
    "ARF"  = "#7b4173"
  )) +
  theme_bw() +
  theme(
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"), 
    legend.position = "none"
  ) +
  labs(
    x = "Facility",
    y = "Faith's Phylogenetic Diversity",
    title = "Faith PD Across Facilities Within ADD Time Groups",
    fill = "Facility"
  )


add_split_faithpd_facility
ggsave(add_split_faithpd_facility, filename = "../05_figures/add_split_faiths_facility.jpeg", device = "jpeg", dpi = 300, width = 8, height = 9)

```


## add_0c + facility + sample_type (question 3)
We know that facility varies across accumulated degree days (add_0c), but it is not the primary variable addressing our research question in this section. Therefore, we include facility in the model as an additive term rather than as an interaction term. Instead, we include sample_type as an interaction with add_0c, because we want to test whether microbial communities from soil and skin change differently over time. We previously examined the effect of facility in an earlier section, so here it is included only as a covariate to account for background variation. Specifically, we are testing whether the time trajectories differ between soil and skin samples. If we instead wanted to compare soil versus skin while simply accounting for time, rather than testing different time trajectories, add_0c would be included as an additive term rather than in an interaction.

Shannon 
```{r}
# Build model
model_add0_facility_sampletype_shannon <- lmer(shannon_entropy ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_shannon)
# look at estimated variance components
summary(model_add0_facility_sampletype_shannon)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_shannon)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_shannon, pairwise ~ sample_type|add0c_split)
```

Evenness
```{r}
# Build model
model_add0_facility_sampletype_evenness <- lmer(pielou_evenness ~ add0c_split*sample_type + facility + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_evenness)
# look at estimated variance components
summary(model_add0_facility_sampletype_evenness)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_evenness)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_evenness, pairwise ~ sample_type|add0c_split)
```

Observed features
```{r}
# Build model
model_add0_facility_sampletype_obfeatures <- lmer(observed_features ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_obfeatures)
# look at estimated variance components
summary(model_add0_facility_sampletype_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_obfeatures)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_obfeatures, pairwise ~ sample_type|add0c_split)
```

Faith's PD
```{r}
# Build model
model_add0_facility_sampletype_faithpd <- lmer(faith_pd ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_faithpd)
# look at estimated variance components
summary(model_add0_facility_sampletype_faithpd)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_faithpd)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_faithpd, pairwise ~ sample_type|add0c_split)
```

## Sample type figure
```{r}
model_add0_facility_sampletype_obfeatures <- lmer(observed_features ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata)  

pvals <- pairs(
  emmeans(model_add0_facility_sampletype_obfeatures, ~ sample_type | add0c_split)
) |> as.data.frame()


pvals_table <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  dplyr::mutate(
    p = round(p.value, 4),
    y.position = max(metadata$observed_features) + 0.1
  )
pvals_table

#only plot the significant associations
pvals_table_significant <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  filter(p.value < 0.05) %>%                     
  mutate(p = round(p.value, 10))
pvals_table_significant

# fix overlapping significance bars
pvals_table_significant <- pvals_table_significant %>%
  add_y_position(
    data = metadata,
    formula = observed_features ~ sample_type,
    fun = "max",
    step.increase = 0.12
  )
pvals_table_significant


# Plot Observed features from skin versus soils add only significant comparisons
add_split_obsFeats_sampleType <- ggplot(metadata,
       aes(x = sample_type, y = observed_features, fill = sample_type)) +
  geom_boxplot() +
  facet_wrap(~ add0c_split) +
  stat_pvalue_manual(
    pvals_table_significant,
    xmin = "group1",
    xmax = "group2",
    label = "p",
    tip.length = 0.01,
    size = 3
  ) +
  scale_fill_manual(values = c(
    "skin" = "#9e003f",
    "soil" = "#b15928")) +
  theme_bw() +
  theme(
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"), 
    legend.position = "none"
  ) +
  labs(
    x = "Sample Type",
    y = "Observed Features",
    title = "Observed Features Within ADD Time Groups",
    fill = "Sample Type"
  )


add_split_obsFeats_sampleType
ggsave(add_split_obsFeats_sampleType, filename = "../05_figures/add_split_obsFeats_sampleType.jpeg", device = "jpeg", dpi = 300, width = 8, height = 9)
```




################################################################################
# Beta diversity
################################################################################

# Load in beta diversity metrics and tidy to add to metadata
(remember this is already our PCOA results so this is already our ordinations). I make separate metadata's for each beta diversity metric, another option would be to rename each PC ex. PC1J for jaccard and add PC1-PC3 for each metric to the metadata. I find it easier to keep everything separate when the metrics have different metadata. The variances extracted from the unzipped PCoA files will not match what you see in the Emperor plot, because Emperor adjusts the variance for plotting. To obtain variances that match the Emperor plot, you need to use qiime2R.
```{r}
# Read in beta div metrics 
bray_curtis <- read_tsv("beta_div/bray_curtis.txt")
jaccard <- read_tsv("beta_div/jaccard.txt")
uw_unifrac <- read_tsv("beta_div/unweighted_unifrac.txt")
w_unifrac <- read_tsv("beta_div/weighted_unifrac.txt")

# Clean metrics
# bray curtis
# get variance
bray_curtis_var <- bray_curtis %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# clean ordination
bray_curtis <- bray_curtis %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_bc <- bray_curtis %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))


# Jaccard
# get variance
jaccard_var <- jaccard %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Clean ordinace
jaccard <- jaccard %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_jaccard <- jaccard %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Unweighted Unifrac
# Get variance
uw_var <- uw_unifrac %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Clean ordinance
uw_unifrac <- uw_unifrac %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_uwunifrac <- uw_unifrac %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Weighted Unifrac
# get variance
wunifrac_var <- w_unifrac %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Tidy ordinace
w_unifrac <- w_unifrac %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(`324`, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_wunifrac <- w_unifrac %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

```
- Now that we've done some data tidying we can finally get to stats for beta diversity, I am going to keep using the add0c split for demonstration purposes. 

- There are two common approaches for performing statistical analyses on beta diversity. One approach is first distances, which we discussed last week. If you want to run a linear mixed-effects model (LME) in R using the first-distances output, you can unzip the first-distances.qza artifact and extract the data file. After importing this file into R, you can use it in your model. In this case, instead of using a PCoA axis (e.g., PC1) as the dependent variable, you would use Distance as the dependent variable in the LME.

-  For the sake of time, we will skip ahead to question 3 and build a model examining sample type. In this model, we account for facility as a covariate and evaluate how beta diversity changes over time. Specifically, we include an interaction between accumulated degree days (add_0c) and sample type to test whether the trajectory of beta diversity over time differs between soil and skin samples. 
## add0c_split + facility + sample_type (question 3)
- If you think back to the PCoA plots we visualized in Emperor, we saw three axes: PC1, PC2, and PC3. We will test each of these axes to determine whether the variance along that axis is explained by our variables. In many microbiome datasets, time often explains a large portion of the variance on PC1, although this is not always the case. 

# Bray Curtis

#PC1
```{r}
# Build model
model_add0_facility_sampletype_bc1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc1)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc1, pairwise ~ sample_type|add0c_split)
```

#PC2
```{r}
# Build model
model_add0_facility_sampletype_bc2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc2)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc2, pairwise ~ sample_type|add0c_split)
```


#PC3
```{r}
# Build model
model_add0_facility_sampletype_bc3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc3)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc3, pairwise ~ sample_type|add0c_split)
```

# Jaccard

PC1
```{r}
# Build model
model_add0_facility_sampletype_j1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j1)
# look at estimated variance components
summary(model_add0_facility_sampletype_j1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j1, pairwise ~ sample_type|add0c_split)
```


PC2
```{r}
# Build model
model_add0_facility_sampletype_j2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j2)
# look at estimated variance components
summary(model_add0_facility_sampletype_j2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j2, pairwise ~ sample_type|add0c_split)
```


PC3
```{r}
# Build model
model_add0_facility_sampletype_j3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j3)
# look at estimated variance components
summary(model_add0_facility_sampletype_j3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j3, pairwise ~ sample_type|add0c_split)
```
# Unweighted Unifrac

PC1
```{r}
# Build model
model_add0_facility_sampletype_uwunifrac1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac1)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac1, pairwise ~ sample_type|add0c_split)
```
PC2
```{r}
# Build model
model_add0_facility_sampletype_uwunifrac2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac2)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac2, pairwise ~ sample_type|add0c_split)
```
PC3
```{r}
# Build model
model_add0_facility_sampletype_uwunifrac3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac3)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac3, pairwise ~ sample_type|add0c_split)
```
# Weighted Unifrac

PC1
```{r}
# Build model
model_add0_facility_sampletype_wunifrac1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac1)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac1, pairwise ~ sample_type|add0c_split)
```
PC2
```{r}
# Build model
model_add0_facility_sampletype_wunifrac2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac2)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac2, pairwise ~ sample_type|add0c_split)
```
PC3
```{r}
# Build model
model_add0_facility_sampletype_wunifrac3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac3)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac3, pairwise ~ sample_type|add0c_split)
```

# Figure

```{r}
# Bray Curtis

# Get min values for placing the p-value label in bottom-left
x_min <- min(metadata_bc$PC1, na.rm = TRUE)
y_min <- min(metadata_bc$PC2, na.rm = TRUE)

# PCoA plot with sample_type coloring, ellipses, and p-value in bottom-left
sampletype_bray_curtis <- ggplot(metadata_bc, aes(x = PC1, y = PC2, color = sample_type)) + 
  geom_point(size = 3) +
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(bray_curtis_var[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(bray_curtis_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("skin" = "#9e003f", "soil" = "#b15928"),
    name = "Sample Type"
  ) +
  labs(title = "Bray-Curtis PCoA") +
  annotate(
    "text",
    x = x_min * 1.15,             
    y = y_min * 1.15,             
    label = "LME; p = 6.618e-12",      # your model p-value, take this from the anova(model_add0_facility_sampletype_bc1) table
    hjust = 0,                     
    vjust = 0,                     
    size = 4,
    fontface = "italic"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold")
  )

sampletype_bray_curtis

# Jaccard
sampletype_jaccard <- ggplot(metadata_jaccard, aes(x = PC1, y = PC2, color = sample_type)) + 
  geom_point(size = 3) +
  theme_bw() + 
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(jaccard_var[1], 1), "%)")) + 
  ylab(paste0("PC2 (", round(jaccard_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("skin" = "#9e003f", "soil" = "#b15928"),
    name = "Sample Type"
  ) +
  labs(title = "Jaccard PCoA") 
# uncomment this section do add a p-value you calculated from your lme
  # annotate("text",
#            x = x_max * 1.2,
#            y = y_max * 1.2,
#            label = "p = 0.0078",
#            hjust = 1, vjust = 1,
#            size = 5, fontface = "italic") +
#   coord_cartesian(clip = "off")

sampletype_jaccard

# Unweighted Unifrac
sampletype_uwunifrac <- ggplot(metadata_uwunifrac, aes(x = PC1, y = PC2, color = sample_type)) + 
  geom_point(size = 3) +
  theme_bw() + 
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(uw_var[1], 1), "%)")) + 
  ylab(paste0("PC2 (", round(uw_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("skin" = "#9e003f", "soil" = "#b15928"),
    name = "Sample Type"
  ) +
  labs(title = "Unweighted Unifrac PCoA") 
# uncomment this section do add a p-value you calculated from your lme
  # annotate("text",
#            x = x_max * 1.2,
#            y = y_max * 1.2,
#            label = "p = 0.0078",
#            hjust = 1, vjust = 1,
#            size = 5, fontface = "italic") +
#   coord_cartesian(clip = "off")

sampletype_uwunifrac

# Weighted Unifrac 
sampletype_wunifrac <- ggplot(metadata_wunifrac, aes(x = PC1, y = PC2, color = sample_type)) + 
  geom_point(size = 3) +
  theme_bw() + 
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(wunifrac_var[1], 1), "%)")) + 
  ylab(paste0("PC2 (", round(wunifrac_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("skin" = "#9e003f", "soil" = "#b15928"),
    name = "Sample Type"
  ) +
  labs(title = "Weighted Unifrac PCoA") 
# uncomment this section do add a p-value you calculated from your lme
  # annotate("text",
#            x = x_max * 1.2,
#            y = y_max * 1.2,
#            label = "p = 0.0078",
#            hjust = 1, vjust = 1,
#            size = 5, fontface = "italic") +
#   coord_cartesian(clip = "off")

sampletype_wunifrac


# Combine metrics plots
beta_div_sample_type <- ggarrange(sampletype_bray_curtis, sampletype_jaccard,
                                  sampletype_uwunifrac, sampletype_wunifrac,
                                  ncol = 2, nrow = 2,
                                  labels = c("A","B","C","D"),
                                  common.legend = TRUE,
                                  legend = "bottom")
beta_div_sample_type

# Save the figure
ggsave(beta_div_sample_type, filename = "../05_figures/beta_div_sample_type.jpeg", device = "jpeg", dpi = 300, width = 10, height = 9)
```

################################################################################
# Taxa bar plots
################################################################################

# Import feature table and taxonomy data from QIIME2 output
```{r}
tabulated_results <- read_tsv("taxonomy/tabulated_results.tsv")
```

# Format, convert taxonomy to relative abundance, and group at taxonomic levels
```{r}
# format data for making a taxa plot
# get rid of #q2:types row
tabulated_results <- tabulated_results[2:nrow(tabulated_results),]

# remove non-numeric columns
# we will filter out Taxon for now so we can calculate relative abundances, but will add it back later
tabulated_results_filt <- tabulated_results %>% 
  select(-all_of(c("Sequence","Confidence","Taxon")))
# we keep id and make it as the row name
tabulated_results_filt <- column_to_rownames(tabulated_results_filt, var = "id")

# calculate relative abundances
tabulated_results_filt[] <- lapply(tabulated_results_filt, function(x) as.numeric(as.character(x)))
rel_abundance <- apply(tabulated_results_filt, 2, function(x) x / sum(x))

# check that each column adds up to 1
colSums(rel_abundance)

# re-add taxonomy to relative abundance table
rel_abundance <- data.frame(rel_abundance, check.names = FALSE)
rel_abundance$id <- rownames(rel_abundance)

tabulated_results_merge <- tabulated_results %>% # create a tabulated results for merging
  select(all_of(c("id","Taxon")))

rel_abundance_taxa <- merge(rel_abundance, tabulated_results_merge, by = "id", all.x = TRUE)

# we now have a table with the per-sample relative abundances of each ASV (along with their taxonomic assignments)

# Taxon column data is messy - let's split the data up into taxonomic ranks
rel_abundance_taxa_split <- rel_abundance_taxa %>% 
  separate(Taxon, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
           sep = ";\\s*",  # splits on semicolon and optional space
           fill = "right", # fills missing columns with NA if some rows are shorter
           remove = TRUE) # remove original "Taxon" column too

# some ranks have a value even though they are unidentified (e.g., g__, s__)
# let's change those to NAs
rel_abundance_taxa_split[rel_abundance_taxa_split == "d__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "p__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "c__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "o__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "f__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "g__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "s__"] <- NA

# create a column with the lowest identified taxonomic level for all ASVs
rel_abundance_taxa_lowest <- rel_abundance_taxa_split %>% 
  rowwise() %>%
  mutate(LowestTaxLevel = {
    tax_levels <- c(Domain, Phylum, Class, Order, Family, Genus, Species)
    last_non_na <- tail(na.omit(tax_levels), 1)
    ifelse(length(last_non_na) > 0, last_non_na, NA)
  }) %>%
  ungroup()

# group by the lowest taxonomic level
rel_abundance_taxa_lowest <- rel_abundance_taxa_lowest %>%
  mutate(LowestTaxLevel = ifelse(is.na(LowestTaxLevel), "Unassigned", LowestTaxLevel)) %>% # convert NAs to "Unassigned"
  group_by(LowestTaxLevel) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# save the file
saveRDS(rel_abundance_taxa_lowest, "taxonomy/rel_abundance_taxa_lowest.RDS")

# group at level 5/Family level
rel_abundance_taxa_family <- rel_abundance_taxa_split %>%
  mutate(Family = ifelse(is.na(Family), "Unassigned", Family)) %>% # convert NAs to "Unassigned"
  group_by(Family) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# save the file
saveRDS(rel_abundance_taxa_family, "taxonomy/rel_abundance_taxa_family.RDS")

# this allows you to have either the lowest-identified level or collapsed taxonomic tables 
# however, these are massive and have >400 samples for this dataset
# because this isn't very conducive to nice taxa bar plots, in the next chunk, we are going to filter to one facility only and make a simple taxa bar plot
```

# Taxa bar plot for STAFS samples
```{r}
# graph at the family level for STAFS samples
# convert to long format
rel_abundance_taxa_family_long <- rel_abundance_taxa_family %>%
  pivot_longer(cols = -Family, names_to = "sample_name", values_to = "relative_abundance")

# assign rare taxa (less than .2 rel abund to other)
rel_abundance_taxa_family_long <- rel_abundance_taxa_family_long %>%
  mutate(rare_taxa = if_else(relative_abundance < 0.2, "Other", Family))

# associate metadata with taxa
metadata <- read_tsv("../03_metadata/metadata.txt")
metadata_filt <- metadata[, which((names(metadata) %in% c("sample_name","sample_type","facility", "add_0c"))==TRUE)]

rel_abundance_taxa_family_long <- rel_abundance_taxa_family_long %>% 
  left_join(metadata_filt, by = "sample_name")

# filter out control samples
rel_abundance_taxa_family_long <- rel_abundance_taxa_family_long[rel_abundance_taxa_family_long$sample_type != "control",]

# filter to just one facility (STAFS)
rel_abundance_taxa_family_long_STAFS <- rel_abundance_taxa_family_long[rel_abundance_taxa_family_long$facility == "STAFS",]

base_colors <- c(
  "#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E",
  "#C06C84", "#A6761D", "#666666", "#7f3b6c", "#355c7d",
  "#b15928", "#9e003f", "#ff7f00", "#8dd3c7", "#bebada",
  "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5",
  "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#a6cee3"
)

# Get unique taxa categories
categories <- unique(rel_abundance_taxa_family_long_STAFS$rare_taxa)

# special colors for 'Other' and 'Unassigned'
special_colors <- c("Other" = "gray", "Unassigned" = "black")
remaining <- setdiff(categories, names(special_colors))

# Assign base colors to remaining taxa
if(length(remaining) > length(base_colors)){
  warning("More taxa than base colors, some colors will be recycled")
}
assigned_colors <- base_colors[seq_along(remaining)]
names(assigned_colors) <- remaining

# Combine special and base colors
colors <- c(special_colors, assigned_colors)

#taxa bar plot, plotted at the Family taxonomic level for STAFS samples
taxa_family_add <- ggplot(rel_abundance_taxa_family_long_STAFS, 
                          aes(x = reorder(sample_name, add_0c), 
                              y = relative_abundance, 
                              fill = rare_taxa)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom",
        axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold")) +
  labs(x = "Sample, sorted by increasing ADD", 
       y = "Relative Abundance", 
       fill = "Family")

taxa_family_add
ggsave(taxa_family_add, filename = "../05_figures/taxa_family_add.jpeg", device = "jpeg", dpi = 300, width = 12, height = 8)

# graph taxa by sample type at the family level for STAFS samples only
rel_abundance_taxa_family_long_STAFS_ordered <- rel_abundance_taxa_family_long_STAFS %>%
  mutate(sample_type = fct_relevel(sample_type, "skin","soil")) %>% 
  arrange(sample_type)

taxa_family_sample_type <- ggplot(rel_abundance_taxa_family_long_STAFS, aes(x = sample_name, y = relative_abundance, fill = rare_taxa)) +
  facet_wrap(~sample_type, scales = "free_x") +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme(legend.position = "bottom",
        axis.text.x = element_blank()) +
  labs(x = "Sample", y = "Relative Abundance", fill = "Family")
taxa_family_sample_type

ggsave(taxa_family_sample_type, filename = "../05_figures/taxa_family_sample_type.jpeg", device = "jpeg", dpi = 300, width = 12, height = 8)
```


# Taxa bar plot for ARF samples at Genus level
```{r}
# group at level 6/Genus level
rel_abundance_taxa_genus <- rel_abundance_taxa_split %>%
  mutate(Genus = ifelse(is.na(Family), "Unassigned", Genus)) %>% # convert NAs to "Unassigned"
  group_by(Genus) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# save the file
saveRDS(rel_abundance_taxa_genus, "taxonomy/rel_abundance_taxa_genus.RDS")

# graph at the genus level for ARF samples
# convert to long format
rel_abundance_taxa_genus_long <- rel_abundance_taxa_genus %>%
  pivot_longer(cols = -Genus, names_to = "sample_name", values_to = "relative_abundance")

# assign rare taxa (less than .2 rel abund to other)
rel_abundance_taxa_genus_long <- rel_abundance_taxa_genus_long %>%
  mutate(rare_taxa = if_else(relative_abundance < 0.2, "Other", Genus))

# associate metadata with taxa
metadata <- read_tsv("../03_metadata/metadata.txt")
metadata_filt <- metadata[, which((names(metadata) %in% c("sample_name","sample_type","facility", "add_0c"))==TRUE)]

rel_abundance_taxa_genus_long <- rel_abundance_taxa_genus_long %>% 
  left_join(metadata_filt, by = "sample_name")

# filter out control samples
rel_abundance_taxa_genus_long <- rel_abundance_taxa_genus_long[rel_abundance_taxa_genus_long$sample_type != "control",]

# filter to just one facility (STAFS)
rel_abundance_taxa_genus_long_ARF <- rel_abundance_taxa_genus_long[rel_abundance_taxa_genus_long$facility == "ARF",]

base_colors <- c(
  "#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E",
  "#C06C84", "#A6761D", "#666666", "#7f3b6c", "#355c7d",
  "#b15928", "#9e003f", "#ff7f00", "#8dd3c7", "#bebada",
  "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5",
  "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#a6cee3"
)

# Get unique taxa categories
categories <- unique(rel_abundance_taxa_genus_long_ARF$rare_taxa)

# special colors for 'Other' and 'Unassigned'
special_colors <- c("Other" = "gray", "Unassigned" = "black")
remaining <- setdiff(categories, names(special_colors))

# Assign base colors to remaining taxa
if(length(remaining) > length(base_colors)){
  warning("More taxa than base colors, some colors will be recycled")
}
assigned_colors <- base_colors[seq_along(remaining)]
names(assigned_colors) <- remaining

# Combine special and base colors
colors <- c(special_colors, assigned_colors)

#taxa bar plot, plotted at the Family taxonomic level for STAFS samples
taxa_genus_add_arf <- ggplot(rel_abundance_taxa_genus_long_ARF, 
                          aes(x = reorder(sample_name, add_0c), 
                              y = relative_abundance, 
                              fill = rare_taxa)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom",
        axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold")) +
  labs(x = "Sample, sorted by increasing ADD", 
       y = "Relative Abundance", 
       fill = "Genus")

taxa_genus_add_arf
ggsave(taxa_genus_add_arf, filename = "../05_figures/taxa_genus_add.jpeg", device = "jpeg", dpi = 300, width = 12, height = 8)

# graph taxa by sample type at the family level for STAFS samples only
rel_abundance_taxa_genus_long_ARF_ordered <- rel_abundance_taxa_genus_long_ARF %>%
  mutate(sample_type = fct_relevel(sample_type, "skin","soil")) %>% 
  arrange(sample_type)

taxa_genus_sample_type_arf <- ggplot(rel_abundance_taxa_genus_long_ARF, aes(x = sample_name, y = relative_abundance, fill = rare_taxa)) +
  facet_wrap(~sample_type, scales = "free_x") +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme(legend.position = "bottom",
        axis.text.x = element_blank()) +
  labs(x = "Sample", y = "Relative Abundance", fill = "Family")
taxa_genus_sample_type_arf

ggsave(taxa_genus_sample_type_arf, filename = "../05_figures/taxa_genus_sample_type_arf.jpeg", device = "jpeg", dpi = 300, width = 12, height = 8)
```








